---
title: "MapleModeling"
author: "Trinity Tobin"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
```{r}
#libraries needed for the project
#any uncommon libraries will include a blurb about their purpose
library(terra)  ##allows me to work with raster data
library(dplyr)
library(readr)
library(ggplot2)
library(corrplot)
library(tidyr)

```



## STEP 1: LOADING IN MY DATA

#loading in my World Clim variables 
#chosen variables: bio1, bio4, bio5, bio12, bio15, bio16 
```{r}
#setting my path to my WorldClimVariables folder 
path <- "C:/Users/trini/Desktop/WorldClimVariables/"

#loading in the my six chosen variables from the 19 
bio1  <- rast(paste0(path, "wc2.1_10m_bio_1.tif"))
bio4  <- rast(paste0(path, "wc2.1_10m_bio_4.tif"))
bio5  <- rast(paste0(path, "wc2.1_10m_bio_5.tif"))
bio12 <- rast(paste0(path, "wc2.1_10m_bio_12.tif"))
bio15 <- rast(paste0(path, "wc2.1_10m_bio_15.tif"))
bio16 <- rast(paste0(path, "wc2.1_10m_bio_16.tif"))

#combining all my variables into a stack (object) so they can be used 
climate_stack <- c(bio1, bio4, bio5, bio12, bio15, bio16)

#checking the stack
plot(climate_stack)
```

```{r}
#restricting my data to North America
#this code cuts off my map geographically using mins and maxs for latitude and longitude to create a bounding box
ext_na <- ext(-170, -50, 10, 80)

climate_na <- crop(climate_stack, ext_na)

plot(climate_na)
```
#loading in my GBIF data
#Beforehand in Excel, the data was separated vis a tab delimiter to give it a better (cleaner) starting point in R. I had originally tried to clean it in R but it became too much of a hassle so I fixed it in Excel
```{r}
sugar_clean <- read_csv("C:/Users/trini/Desktop/GBIF_Acer.csv")

#convert my GBIF data to spatial data
sugar_points <- vect(
  sugar_clean,
  geom = c("decimalLongitude", "decimalLatitude"),
  crs = "EPSG:4326"
)

#renaming my WorldClim variable columns 
names(climate_na) <- c("BIO1", "BIO4", "BIO5", "BIO12", "BIO15", "BIO16")

#THE CODE THAT COMBINES THE GBIF AND THE WORLD CLIM
climate_extract <- extract(climate_na, sugar_points)

sugar_climate <- bind_cols(sugar_clean, climate_extract[, -1])
head(sugar_climate)
```
#from this point, it is clear that there is a lot of cleaning and processing that needs to be done before Exploratory Data Analysis (because the majority of these columns have no use to my analysis) but the datasets are together.


```{r}
colnames(sugar_climate)
```
# STEP 2: FIXING MY DATA FOR EXPLORATORY DATA ANALYSIS (EDA)

#creating a table that is a subset of my data to fix the World Clim variable issue 
```{r}
#extracting columns for the table
sugar_coords <- sugar_clean %>%
  transmute(
    species = as.character(species),
    scientificName = as.character(scientificName),
    decimalLatitude = as.numeric(decimalLatitude),
    decimalLongitude = as.numeric(decimalLongitude),
    countryCode = as.character(countryCode),
    stateProvince = as.character(stateProvince)
  )

#removing coordinates that are not within the bounding box (impossible within the confines of our geographic range)
sugar_coords <- sugar_coords %>%
  filter(decimalLatitude >= -90, decimalLatitude <= 90) %>%
  filter(decimalLongitude >= -180, decimalLongitude <= 180)

#removing missing coordinates 
#sugar_coords <- sugar_coords %>%
#  filter(!is.na(decimalLatitude), !is.na(decimalLongitude))

#keeping the North American coordinates 
sugar_coords <- sugar_coords %>%
  filter(decimalLatitude >= 10, decimalLatitude <= 80) %>%
  filter(decimalLongitude >= -170, decimalLongitude <= -50)

#checking the ranges 
sugar_coords <- sugar_coords %>%
  filter(decimalLatitude >= 10, decimalLatitude <= 80) %>%
  filter(decimalLongitude >= -170, decimalLongitude <= -50)


#NOTE: I did not remove any missing (or duplicate values, although I don't plan to remove duplicate values, as we are targeting abundance) yet, as I feel those can be dealt with during data preprocessing. The cleaning that I did was only out of necessity to join the data sets. I did create the code to do so and it will be here. 

head(sugar_coords)
```

```{r}
#remaking the points object from earlier (line 61) but within my new table
sugar_points2 <- vect(
  sugar_coords,
  geom = c("decimalLongitude", "decimalLatitude"),
  crs = "EPSG:4326"
)

#checking that the points are in the raster
ext(climate_na)
ext(sugar_points2)

plot(climate_na[[1]])
points(sugar_points2, col = "red", pch = 20, cex = 0.4)
```
```{r}
#reruning my extraction
names(climate_na) <- c("BIO1", "BIO4", "BIO5", "BIO12", "BIO15", "BIO16")

climate_extract2 <- extract(climate_na, sugar_points2)

colSums(is.na(climate_extract2))
head(climate_extract2)

#combining the climate values with my cleaned table 
sugar_climate2 <- bind_cols(sugar_coords, climate_extract2[, -1])

head(sugar_climate2)
summary(sugar_climate2[, c("BIO1", "BIO4", "BIO5", "BIO12", "BIO15", "BIO16")])

```

# STEP 3: EXPLORATORY DATA ANALYSIS


FOR THE REQUIREMENTS OF THE EDA:
Table 1: sugar_climate2 (ABOVE)

Table 2:
```{r}
#summary statistics table of the WorldClim variables
bio_summary_table <- sugar_climate2 %>%
  summarise(
    BIO1_mean = mean(BIO1, na.rm = TRUE),
    BIO1_sd   = sd(BIO1, na.rm = TRUE),
    
    BIO4_mean = mean(BIO4, na.rm = TRUE),
    BIO4_sd   = sd(BIO4, na.rm = TRUE),
    
    BIO5_mean = mean(BIO5, na.rm = TRUE),
    BIO5_sd   = sd(BIO5, na.rm = TRUE),
    
    BIO12_mean = mean(BIO12, na.rm = TRUE),
    BIO12_sd   = sd(BIO12, na.rm = TRUE),
    
    BIO15_mean = mean(BIO15, na.rm = TRUE),
    BIO15_sd   = sd(BIO15, na.rm = TRUE),
    
    BIO16_mean = mean(BIO16, na.rm = TRUE),
    BIO16_sd   = sd(BIO16, na.rm = TRUE)
  )

bio_summary_table

```


Graph 1: Histogram of Temperature
Graph 2: Histogram of Precipitation
```{r}
long_data <- sugar_climate2 %>%
  select(BIO1, BIO4, BIO5, BIO12, BIO15, BIO16) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Value")

ggplot(long_data, aes(x = Value)) +
  geom_histogram(bins = 30) +
  facet_wrap(~Variable, scales = "free") +
  labs(title = "Distributions of Climate Variables")
```

Graph 3: Scatterplot of BIO1 V BIO12
```{r}
ggplot(sugar_climate2, aes(x = BIO1, y = BIO12)) +
  geom_point(alpha = 0.4) +
  labs(title = "Temperature vs Precipitation")
```


Graph 4: Correlation Heatmap
```{r}
climate_data <- sugar_climate2 %>%
  select(BIO1, BIO4, BIO5, BIO12, BIO15, BIO16)

cor_matrix <- cor(climate_data, use = "complete.obs")

corrplot(cor_matrix, method = "color")
```

Graph 5: Geographic Map of Occurrences
```{r}
ggplot(sugar_climate2, aes(x = decimalLongitude, y = decimalLatitude)) +
  geom_point(alpha = 0.3) +
  labs(title = "Geographic Distribution of Sugar Maple")
```


Graph 6: Boxplot of BIO1
```{r}
ggplot(sugar_climate2, aes(y = BIO1)) +
  geom_boxplot() +
  labs(title = "Distribution of Annual Temperature (BIO1)")
```


Graph 7: Density map of sugar maple occurrences 
```{r}
ggplot(sugar_climate2, aes(x = decimalLongitude, y = decimalLatitude)) +
  geom_bin2d() +
  labs(title = "Density of Sugar Maple Occurrences")
```
```{r}
ggplot(sugar_climate2, aes(x = decimalLongitude, y = decimalLatitude)) +
  geom_hex() +
  labs(title = "Hexbin Density of Sugar Maple Occurrences")
```
```{r}
ggplot(sugar_climate2, aes(x = decimalLongitude, y = decimalLatitude, color = BIO1)) +
  geom_point(alpha = 0.5) +
  labs(title = "Sugar Maple Occurrences Colored by Temperature (BIO1)")
```




GOALS FOR TOMORROW:

#1: join your sugar_climate2 dataset with the main data set (optional; ask Dr.Geist)
#2: finish coding all of your graphs and tables (Part 1)
#3: finish summary table (add min, max, median and fix formating)


GOALS FOR THE  NEXT DAY: 
#1: finish the figures and captions for your graphs and tables



