---
title: "MapleModeling"
author: "Trinity Tobin"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
```{r}
#libraries needed for the project
#any uncommon libraries will include a blurb about their purpose
library(terra)  ##allows me to work with raster data
library(dplyr)
library(readr)
library(ggplot2)
library(corrplot) ##for heat map
library(tidyr)
library(psych)  ##for summary stats table
library(flextable)  ##helps me take my tables from here to word (they are large)
library(knitr)
library(caret)  ##for my correlation stuff in preprocessing 

```



## STEP 1: LOADING IN MY DATA

#loading in my World Clim variables 
#chosen variables: bio1, bio4, bio5, bio12, bio15, bio16 
```{r}
#setting my path to my WorldClimVariables folder 
path <- "C:/Users/trini/Desktop/WorldClimVariables/"

#loading in the my six chosen variables from the 19 
bio1  <- rast(paste0(path, "wc2.1_10m_bio_1.tif"))
bio4  <- rast(paste0(path, "wc2.1_10m_bio_4.tif"))
bio5  <- rast(paste0(path, "wc2.1_10m_bio_5.tif"))
bio12 <- rast(paste0(path, "wc2.1_10m_bio_12.tif"))
bio15 <- rast(paste0(path, "wc2.1_10m_bio_15.tif"))
bio16 <- rast(paste0(path, "wc2.1_10m_bio_16.tif"))

#combining all my variables into a stack (object) so they can be used 
climate_stack <- c(bio1, bio4, bio5, bio12, bio15, bio16)

#checking the stack
plot(climate_stack)
```

```{r, fig.width=10, fig.height=6, echo=FALSE, warning=FALSE, message=FALSE}
#restricting my data to North America
#this code cuts off my map geographically using mins and maxs for latitude and longitude to create a bounding box
ext_na <- ext(-170, -50, 10, 80)

climate_na <- crop(climate_stack, ext_na)

names(climate_na) <- c(
  "BIO1",
  "BIO4",
  "BIO5",
  "BIO12",
  "BIO15",
  "BIO16"
)

#make room for one outer title
par(oma = c(0, 0, 4, 0))

#plot the 6 layers
plot(climate_na)

#add one overall title
mtext(
  "Figure 1. Spatial Distribution of Selected WorldClim Variables Across North America",
  side = 3,
  outer = TRUE,
  cex = 1.2,
  font = 2,
  line = 1
)

```
#loading in my GBIF data
#Beforehand in Excel, the data was separated vis a tab delimiter to give it a better (cleaner) starting point in R. I had originally tried to clean it in R but it became too much of a hassle so I fixed it in Excel
```{r}
sugar_clean <- read_csv("C:/Users/trini/Desktop/GBIF_Acer.csv")

#convert my GBIF data to spatial data
sugar_points <- vect(
  sugar_clean,
  geom = c("decimalLongitude", "decimalLatitude"),
  crs = "EPSG:4326"
)

#renaming my WorldClim variable columns 
names(climate_na) <- c("BIO1", "BIO4", "BIO5", "BIO12", "BIO15", "BIO16")

#THE CODE THAT COMBINES THE GBIF AND THE WORLD CLIM
climate_extract <- terra::extract(climate_na, sugar_points)

sugar_climate <- bind_cols(sugar_clean, climate_extract[, -1])
head(sugar_climate)
```
#from this point, it is clear that there is a lot of cleaning and processing that needs to be done before Exploratory Data Analysis (because the majority of these columns have no use to my analysis) but the datasets are together.


```{r}
colnames(sugar_climate)
```
# STEP 2: FIXING MY DATA FOR EXPLORATORY DATA ANALYSIS (EDA)

#creating a table that is a subset of my data to fix the World Clim variable issue 
```{r}
#extracting columns for the table
sugar_coords <- sugar_clean %>%
  transmute(
    species = as.character(species),
    scientificName = as.character(scientificName),
    decimalLatitude = as.numeric(decimalLatitude),
    decimalLongitude = as.numeric(decimalLongitude),
    countryCode = as.character(countryCode),
    stateProvince = as.character(stateProvince),
     year = as.numeric(year)
  )

#removing coordinates that are not within the bounding box (impossible within the confines of our geographic range)
sugar_coords <- sugar_coords %>%
  filter(decimalLatitude >= -90, decimalLatitude <= 90) %>%
  filter(decimalLongitude >= -180, decimalLongitude <= 180)

#removing missing coordinates 
#sugar_coords <- sugar_coords %>%
#  filter(!is.na(decimalLatitude), !is.na(decimalLongitude))

#keeping the North American coordinates 
sugar_coords <- sugar_coords %>%
  filter(decimalLatitude >= 10, decimalLatitude <= 80) %>%
  filter(decimalLongitude >= -170, decimalLongitude <= -50)

#checking the ranges 
sugar_coords <- sugar_coords %>%
  filter(decimalLatitude >= 10, decimalLatitude <= 80) %>%
  filter(decimalLongitude >= -170, decimalLongitude <= -50)


#NOTE: I did not remove any missing (or duplicate values, although I don't plan to remove duplicate values, as we are targeting abundance) yet, as I feel those can be dealt with during data preprocessing. The cleaning that I did was only out of necessity to join the data sets. I did create the code to do so and it will be here. 

head(sugar_coords)
```

```{r, fig.width=10, fig.height=6, echo=FALSE, warning=FALSE, message=FALSE}
#remaking the points object from earlier (line 61) but within my new table
sugar_points2 <- vect(
  sugar_coords,
  geom = c("decimalLongitude", "decimalLatitude"),
  crs = "EPSG:4326"
)

#checking that the points are in the raster
ext(climate_na)
ext(sugar_points2)

# bigger outer margins
par(mar = c(6, 5, 5, 9), xpd = NA)

plot(
  climate_na[["BIO1"]],
  main = "Figure 2. Sugar Maple Occurrences Over BIO1 (Annual Mean Temperature)",
  xlab = "",
  ylab = "Latitude",
  axes = TRUE,
  plg = list(
    x = 1.05,
    title = "Annual Mean\nTemperature (°C)"
  )
)

points(
  sugar_points2,
  col = "black",
  pch = 16,
  cex = 0.6
)

#not get cut off
mtext("Longitude", side = 1, line = 3)

```
```{r}
#reruning my extraction
names(climate_na) <- c("BIO1", "BIO4", "BIO5", "BIO12", "BIO15", "BIO16")

climate_extract2 <- terra::extract(climate_na, sugar_points2)

colSums(is.na(climate_extract2))
head(climate_extract2)

#combining the climate values with my cleaned table 
sugar_climate2 <- bind_cols(sugar_coords, climate_extract2[, -1])

head(sugar_climate2)
summary(sugar_climate2[, c("BIO1", "BIO4", "BIO5", "BIO12", "BIO15", "BIO16")])

```

# STEP 3: EXPLORATORY DATA ANALYSIS


FOR THE REQUIREMENTS OF THE EDA:
Table 1: sugar_climate2 (ABOVE)

Table 2:
```{r}
#summary statistics table of the WorldClim variables
bio_summary_table <- describe(
  sugar_climate2 %>%
    select(BIO1, BIO4, BIO5, BIO12, BIO15, BIO16)
)

kable(
  bio_summary_table,
  caption = "Table 1. Summary Statistics for Selected WorldClim Variables",
  digits = 2
)


```


Graph 1: Histogram of Temperature
Graph 2: Histogram of Precipitation

#rather than make separate histograms, I made one figure with each BIO variable represented as a histogram
```{r}
long_data <- sugar_climate2 %>%
  select(BIO1, BIO4, BIO5, BIO12, BIO15, BIO16) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Value")


theme_report <- theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

bio_long <- sugar_climate2 %>%
  select(BIO1, BIO4, BIO5, BIO12, BIO15, BIO16) %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Value") %>%
  mutate(
    Variable = factor(
      Variable,
      levels = c("BIO1", "BIO4", "BIO5", "BIO12", "BIO15", "BIO16")
    )
  )

ggplot(bio_long, aes(x = Value, fill = Variable)) +
  geom_histogram(bins = 30, color = "black", alpha = 0.8) +
  facet_wrap(~ Variable, scales = "free") +
  labs(
    title = "Figure 3. Distributions of Selected WorldClim Variables",
    x = "Value",
    y = "Frequency"
  ) +
  theme_report

```

Graph 3: Scatterplot of BIO1 V BIO12
```{r}
ggplot(sugar_climate2, aes(x = BIO1, y = BIO12)) +
  geom_point(alpha = 0.4) +
  labs(title = "Temperature vs Precipitation") +
  theme_report
```


Graph 4: Correlation Heatmap
```{r}
library(corrplot)
library(dplyr)

climate_data <- sugar_climate2 %>%
  select(BIO1, BIO4, BIO5, BIO12, BIO15, BIO16)

cor_matrix <- cor(climate_data, use = "complete.obs")

# increase margins for spacing
par(mar = c(1, 1, 3, 1))

corrplot(
  cor_matrix,
  method = "color",
  type = "upper",
  col = colorRampPalette(c("#4575b4", "white", "#d73027"))(200),
  tl.col = "black",
  tl.cex = 1,
  tl.srt = 30,             # less extreme angle
  addCoef.col = "black",
  number.cex = 0.8,
  cl.cex = 0.8
)

```

Graph 5: Geographic Map of Occurrences
```{r}
ggplot(sugar_climate2, aes(x = decimalLongitude, y = decimalLatitude)) +
  geom_point(alpha = 0.3) +
  labs(title = "Geographic Distribution of Sugar Maple") +
  theme_report
```


Graph 6: Boxplot of BIO1
```{r}
ggplot(sugar_climate2, aes(y = BIO1)) +
  geom_boxplot() +
  labs(title = "Distribution of Annual Temperature (BIO1)") +
  theme_report
```


Graph 7: Density map of sugar maple occurrences 
```{r}
ggplot(sugar_climate2, aes(x = decimalLongitude, y = decimalLatitude)) +
  geom_hex() +
  labs(title = "Hexbin Density of Sugar Maple Occurrences") +
  theme_report
```

```{r}
ggplot(sugar_climate2, aes(x = decimalLongitude, y = decimalLatitude, color = BIO1)) +
  geom_point(alpha = 0.5) +
  labs(title = "Sugar Maple Occurrences Colored by Temperature (BIO1)") +
  theme_report
```

Graph 8: Line Graph of year v. precipitation
```{r}
precip_by_year <- sugar_climate2 %>%
  filter(!is.na(year)) %>%
  group_by(year) %>%
  summarise(mean_precip = mean(BIO12, na.rm = TRUE))

ggplot(precip_by_year, aes(x = year, y = mean_precip)) +
  geom_line() +
  labs(
    title = "Figure 5. Mean Annual Precipitation by Observation Year",
    x = "Year",
    y = "Mean Precipitation (BIO12)"
  ) +
  theme_report
```

Graph 9: Line Graph of year v. temperature 
```{r}
temp_by_year <- sugar_climate2 %>%
  filter(!is.na(year)) %>%
  group_by(year) %>%
  summarise(mean_temp = mean(BIO1, na.rm = TRUE))

ggplot(temp_by_year, aes(x = year, y = mean_temp)) +
  geom_line() +
  labs(
    title = "Figure 6. Mean Annual Temperature by Observation Year",
    x = "Year",
    y = "Mean Temperature (BIO1)"
  ) +
  theme_report
```

Formatting the line graphs together
```{r}
precip_by_year <- sugar_climate2 %>%
  filter(!is.na(year), year >= 1950) %>%
  group_by(year) %>%
  summarise(mean_precip = mean(BIO12, na.rm = TRUE))

temp_by_year <- sugar_climate2 %>%
  filter(!is.na(year), year >= 1950) %>%
  group_by(year) %>%
  summarise(mean_temp = mean(BIO1, na.rm = TRUE))

#precipitation
ggplot(precip_by_year, aes(x = year, y = mean_precip)) +
  geom_line(color = "#1f78b4", alpha = 0.6) +
  geom_smooth(color = "#e31a1c", se = FALSE, size = 1.2) +
  labs(
    title = "Figure 5. Mean Annual Precipitation by Observation Year",
    x = "Year",
    y = "Mean Precipitation (BIO12)"
  ) +
  theme_report

#temperature
ggplot(temp_by_year, aes(x = year, y = mean_temp)) +
  geom_line(color = "#33a02c", alpha = 0.6) +
  geom_smooth(color = "#e31a1c", se = FALSE, size = 1.2) +
  labs(
    title = "Figure 6. Mean Annual Temperature by Observation Year",
    x = "Year",
    y = "Mean Temperature (BIO1)"
  ) +
  theme_report
```

Table 3: Summary of GBIF spatial distribution
```{r}
#calculating abundance as an average per state
state_abundance <- sugar_climate2 %>%
  group_by(stateProvince) %>%
  summarise(count = n())

#building the table 
distribution_table <- sugar_climate2 %>%
  summarise(
    total_abundance = n(),
    avg_abundance = mean(state_abundance$count),
    min_latitude = min(decimalLatitude, na.rm = TRUE),
    max_latitude = max(decimalLatitude, na.rm = TRUE),
    min_longitude = min(decimalLongitude, na.rm = TRUE),
    max_longitude = max(decimalLongitude, na.rm = TRUE)
  )

distribution_table

kable(
  distribution_table,
  caption = "Table 2. Spatial Distribution Summary of Sugar Maple Occurrences",
  digits = 2
)

```



















## STEP 4:DATA PREPROCESSING

General Steps
1. Train-test split ratio
2. Encode categorical variables 
3. Impute missing values 
4. Perform arithmetic transformations
5. Normalize, center, and/or scale

```{r}
#WorldClim -> importing the remaining 13 variables and assessing missing data 
#Adding the remaining varaibles won't overwhelm my data (it's only 13) and will create a more robust analysis and a more robust model.

path <- "C:/Users/trini/Desktop/WorldClimVariables/"

#loading in all 19 WorldClim variables
bio_files <- paste0(path, "wc2.1_10m_bio_", 1:19, ".tif")

#combining all my variables into a stack (object) so they can be used
climate_stack_ALL <- rast(bio_files)
names(climate_stack_ALL) <- paste0("BIO", 1:19)

#checking the stack
plot(climate_stack_ALL)
```

```{r}
#removing rows with missing/invalid coordinates before making points
#WHY: This was added after troubleshooting combining the data sets and it allows me to have values for the BIO WorldClim variables.

sugar_clean <- sugar_clean %>%
  filter(
    !is.na(decimalLongitude),
    !is.na(decimalLatitude),
    decimalLongitude >= -180, decimalLongitude <= 180,
    decimalLatitude >= -90, decimalLatitude <= 90
  )
```


```{r}

#making lat long points from cleaned GBIF data
sugar_points <- vect(
  sugar_clean,
  geom = c("decimalLongitude", "decimalLatitude"),
  crs = "EPSG:4326"
)

#matching CRS
if (!same.crs(sugar_points, climate_stack_ALL)) {
  sugar_points <- project(sugar_points, crs(climate_stack_ALL))
}

#extracting all 19 BIO values to the GBIF points
climate_extract_19 <- terra::extract(climate_stack_ALL, sugar_points)

#combining GBIF + extracted climate values
sugar_climate_clean <- bind_cols(sugar_clean, climate_extract_19[, -1])

```


```{r}
#cleaning my data 
#GBIF -> getting rid of unnecessary values and assessing missing data 
#The focus of my project is on abundance and time, and there are lots of variables related to the actual data collection that are not necessary to include in my model.

#colnames(sugar_climate)


#There are a lot of variables to remove so we'll do this in a step-wise way so I don't get rid of anything I truly need and so that I can find it quickly and add it back if I need to

sugar_climate_clean <- sugar_climate_clean %>%
  select(-c(
    #Identifiers
    #We don't need thesae because they aren't gonna help predict anything and they could cause overfitting
    
    gbifID, datasetKey, occurrenceID, taxonKey, speciesKey,
    institutionCode, collectionCode, catalogNumber,
    recordNumber, identifiedBy, recordedBy,
    license, rightsHolder,
    
    #Taxonomic Hierachy 
    #We don't need these because my project is only targeting one species (their inclusion is also redundant) and their inclusion could increase multicollinearity and noise in the models 
    
    kingdom, phylum, class, order, family, taxonRank,
    
    #species naming
    #We don't need to include the names multiple times and there's too much text
    
    scientificName, verbatimScientificName,
    verbatimScientificNameAuthorship, infraspecificEpithet,
    
    #Location text
    #We are using lat and long for the points (but might include back later on), which are more precise for the model.
    
    countryCode, stateProvince, locality,
    
    #Data collection stuff
    #We don't need because there is no biological relevance to my project
    
    publishingOrgKey, basisOfRecord,
    typeStatus, issue, establishmentMeans, lastInterpreted, mediaType,
    
    #Precision / uncertainty
    #We don't need for predictive modeling (but could be used for data quality analysis, which could be a good metric to include).
    
    coordinateUncertaintyInMeters, coordinatePrecision,
    elevationAccuracy, depthAccuracy,
    
    #Depth 
    depth
  ))

colnames(sugar_climate_clean)

```

```{r}
#checking my newly combined data before proceeding
head(sugar_climate_clean)
```


#ENCODING VARIABLES
```{r}
#1. individualCount
#WHY: for occurrence counts for my abundance because each row counts as an occurrence
sugar_climate_clean <- sugar_climate_clean %>%
  mutate(individualCount = ifelse(is.na(individualCount), 1, individualCount))

```


#CREATING MY LONGITUDINAL SLICES 
```{r}
#WHY: for my model, for my time series, and for my abundance measurements 
#HOW: creating bins 

#I made three different versions of the code, as right now I am not sure how big I want my bins to be 

#10-degree bins
#sugar_climate_clean <- sugar_climate_clean %>%
#  mutate(longitude_band = floor(decimalLongitude / 10) * 10)

#5-degree bins
sugar_climate_clean <- sugar_climate_clean %>%
  mutate(longitude_band = floor(decimalLongitude / 5) * 5)

#2-degree bins 
#sugar_climate_clean <- sugar_climate_clean %>%
#  mutate(longitude_band = floor(decimalLongitude / 2) * 2)

```


#CREATING MY ABUNDANCE VARIABLE
```{r}
#Abundance as a variable in EDA was defined as the sum of occurrences in a state. While that worked for then, I want to make that stronger and calculate the abundance as the sum of occurrences/Individual counts. Additionally, I can do a sum abundance measurement and a relative abundance measurement. 

#WHY I'M NOT USING STATE ANYMORE: sampling bias. Ex.: people in New York are more likely to sample that people in South Carolina (something like that but mroe fleshed out).

#this code creates the modeling table (essentially, it is aggregating the data by both  longitude_band and year into a useable format for future modeling)

abundance_data <- sugar_climate_clean %>%
  group_by(longitude_band, year) %>%
  summarise(
    abundance = sum(individualCount, na.rm = TRUE),
    occurrences = n(),
    decimalLatitude = mean(decimalLatitude, na.rm = TRUE),
    decimalLongitude = mean(decimalLongitude, na.rm = TRUE),
    elevation = mean(elevation, na.rm = TRUE),
    BIO1 = mean(BIO1, na.rm = TRUE),
    BIO2 = mean(BIO2, na.rm = TRUE),
    BIO3 = mean(BIO3, na.rm = TRUE),
    BIO4 = mean(BIO4, na.rm = TRUE),
    BIO5 = mean(BIO5, na.rm = TRUE),
    BIO6 = mean(BIO6, na.rm = TRUE),
    BIO7 = mean(BIO7, na.rm = TRUE),
    BIO8 = mean(BIO8, na.rm = TRUE),
    BIO9 = mean(BIO9, na.rm = TRUE),
    BIO10 = mean(BIO10, na.rm = TRUE),
    BIO11 = mean(BIO11, na.rm = TRUE),
    BIO12 = mean(BIO12, na.rm = TRUE),
    BIO13 = mean(BIO13, na.rm = TRUE),
    BIO14 = mean(BIO14, na.rm = TRUE),
    BIO15 = mean(BIO15, na.rm = TRUE),
    BIO16 = mean(BIO16, na.rm = TRUE),
    BIO17 = mean(BIO17, na.rm = TRUE),
    BIO18 = mean(BIO18, na.rm = TRUE),
    BIO19 = mean(BIO19, na.rm = TRUE),
    .groups = "drop"
  )

head(abundance_data)



#1. add abundance as a target variable in the data set
#2. figure out how to segment abundance based on longitudinal lines
#3. log transformation of abundance (helps model and skew of variable)
```


#CREATING RELATIVE ABUNDANCE
```{r}
abundance_data <- abundance_data %>%
  group_by(year) %>%
  mutate(relative_abundance = abundance / sum(abundance, na.rm = TRUE)) %>%
  ungroup()
```


#LOG TRANSFORMING MY ABUNDANCE VARIABLE
```{r}
abundance_data <- abundance_data %>%
  mutate(log_abundance = log1p(abundance))
```




PLAN FOR 4/12:
1. Create a time-lagged abundance variable within each longitude band
WHY: for the temporal aspect of my model, to model abundance over time
```{r}
#creating time-lagged abundance variables within each longitude band

#What is actually happening here is that


abundance_data <- abundance_data %>%
  arrange(longitude_band, year) %>%
  group_by(longitude_band) %>%
  mutate(
    lag_abundance = lag(abundance, 1),
    lag_log_abundance = lag(log_abundance, 1),
    lag_relative_abundance = lag(relative_abundance, 1)
  ) %>%
  ungroup()

head(abundance_data)

```

2. Generate additional derived variables such as climate ranges and interactions
WHY: to have more that can be analyzed in the model, to provide more model insights
I did this by combing mainly the BIO variables to provide more data regarding the preferred ranges of the species
```{r}
#creating more BIO variables
abundance_data <- abundance_data %>%
  mutate(
    temp_range = BIO5 - BIO6,
    temp_wet_quarter_diff = BIO8 - BIO9,
    precip_range = BIO16 - BIO14,
    precip_seasonality = BIO15,
    year_centered = year - mean(year, na.rm = TRUE)
  )

head(abundance_data)
summary(abundance_data[, c("temp_range", "temp_wet_quarter_diff", "precip_range", 
                           "precip_seasonality", "year_centered")])


#creating more GBIF variables

#WHY: I tried creating latitudinal slices as well (in case I wanted to use those instead of or with my longitudinal slices to create a grid. Plus the code was the same). 

#abundance_data <- abundance_data %>%
#  mutate(latitude_band = floor(decimalLatitude / 5) * 5)




#WHY: this variable can measure how spread out my occurrences are and can allow me to look at the range of the species (as well as if it is expanding or being restricted)
abundance_data <- abundance_data %>%
  mutate(
    center_lat = mean(decimalLatitude, na.rm = TRUE),
    center_lon = mean(decimalLongitude, na.rm = TRUE),
    distance_from_center = sqrt((decimalLatitude - center_lat)^2 +
                                (decimalLongitude - center_lon)^2)
  )


#WHY: this variable will help me look at changes in growth/decline patterns and serve as a predictor 
abundance_data <- abundance_data %>%
  arrange(longitude_band, year) %>%
  group_by(longitude_band) %>%
  mutate(abundance_change = abundance - lag(abundance)) %>%
  ungroup()


#WHY: this variable controls for the sampling effort and relps reduce bias 
abundance_data <- abundance_data %>%
  mutate(density = abundance / occurrences)

```

3. Missing values in the dataset
WHY: missing values will impact the models 
```{r}
#checking for missing 
colSums(is.na(abundance_data))
```
```{r}
#how I am dealing with the missing variables 
# 1.Lagging my variable creates a missing value in the data set, and the easiest way to deal with those is to remove them.

#removing rows with missing lag variables
abundance_data <- abundance_data %>%
  filter(
    !is.na(lag_abundance),
    !is.na(lag_log_abundance),
    !is.na(lag_relative_abundance)
  )

#checking missing values
#colSums(is.na(abundance_data))



# 2. There isn't a lot of missing stuff after that so imputation can be done based on the median to fill those points in so the data is comparable. 

#median imputation for remaining numeric missing values
abundance_data <- abundance_data %>%
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), median(., na.rm = TRUE), .)))

#justify
#assess your imputation by calculating percentage (low) and comparing bf and after (prove no shift)

#checking again
colSums(is.na(abundance_data))
```

4. Finalizing the target variable for modeling.
```{r}
# It is important to know in preprocessing which variable is my target. Abundance is my target variables, but we log transformed it before so the new target is log_abundance

# Outlining them now will help me to frame my mindset and my approach to the data.
# The remaining abundance variables ("abundance" and "relative_abundance") will serve as secondary ways to evaluate abundance in my dataset and help with my interpretation.

#my target variable
y <- abundance_data$log_abundance

#my model predictors 
x <- abundance_data %>%
  select(-abundance, -relative_abundance, -log_abundance)
```


5. Working on my longitudinal bands 
```{r}
#WHY: I made the code for 2-, 5-, and 10-year bands so I could have the option to choose but it is better to evaluate how each one performs and choose one to move forward with for my model.

#10 degree
sugar_climate_clean_10 <- sugar_climate_clean %>%
  mutate(longitude_band = floor(decimalLongitude / 10) * 10)

abundance_data_10 <- sugar_climate_clean_10 %>%
  mutate(individualCount = ifelse(is.na(individualCount), 1, individualCount)) %>%
  group_by(longitude_band, year) %>%
  summarise(
    abundance = sum(individualCount, na.rm = TRUE),
    occurrences = n(),
    .groups = "drop"
  )

nrow(abundance_data_10)
summary(abundance_data_10$abundance)


#5 year 
sugar_climate_clean_5 <- sugar_climate_clean %>%
  mutate(longitude_band = floor(decimalLongitude / 5) * 5)

abundance_data_5 <- sugar_climate_clean_5 %>%
  mutate(individualCount = ifelse(is.na(individualCount), 1, individualCount)) %>%
  group_by(longitude_band, year) %>%
  summarise(
    abundance = sum(individualCount, na.rm = TRUE),
    occurrences = n(),
    .groups = "drop"
  )

nrow(abundance_data_5)
summary(abundance_data_5$abundance)


#2 year 
sugar_climate_clean_2 <- sugar_climate_clean %>%
  mutate(longitude_band = floor(decimalLongitude / 2) * 2)

abundance_data_2 <- sugar_climate_clean_2 %>%
  mutate(individualCount = ifelse(is.na(individualCount), 1, individualCount)) %>%
  group_by(longitude_band, year) %>%
  summarise(
    abundance = sum(individualCount, na.rm = TRUE),
    occurrences = n(),
    .groups = "drop"
  )

nrow(abundance_data_2)
summary(abundance_data_2$abundance)


#comparing them with histograms 
hist(abundance_data_10$abundance, main = "10-degree bands", xlab = "Abundance")
hist(abundance_data_5$abundance, main = "5-degree bands", xlab = "Abundance")
hist(abundance_data_2$abundance, main = "2-degree bands", xlab = "Abundance")
```
CONCLUSION: There isn't really a difference in the abundance distribution across the degree segments so I'm probably gonna use 5-year. 



6. Perform a train-test split on the aggregated dataset.
WHY: all models have training and test sets, no biggie 
```{r}
set.seed(123)

n <- nrow(abundance_data)
train_index <- sample(seq_len(n), size = 0.8 * n)

train_data <- abundance_data[train_index, ]
test_data  <- abundance_data[-train_index, ]

#checking sizes
nrow(train_data)
nrow(test_data)
```

7. Impute any remaining missing predictor values
WHY: same reasoning as earlier, will affect model performance
```{r}
#imputing numeric missing values in training data
train_data <- train_data %>%
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), median(., na.rm = TRUE), .)))

#imputing numeric missing values in test data
test_data <- test_data %>%
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), median(., na.rm = TRUE), .)))

```

8. Normalize and scale numeric variables for modeling.
```{r}
#I am scaling everything by log_abundance (my target)

#selecting numeric predictor columns only
numeric_cols <- train_data %>%
  select(where(is.numeric)) %>%
  select(-abundance, -relative_abundance, -log_abundance, -longitude_band) %>%
  colnames()

#storing training means and sds
train_means <- sapply(train_data[, numeric_cols], mean, na.rm = TRUE)
train_sds   <- sapply(train_data[, numeric_cols], sd, na.rm = TRUE)

#scaling training data
train_data[, numeric_cols] <- scale(train_data[, numeric_cols], center = train_means, scale = train_sds)

#scaling test data using training means and sds
test_data[, numeric_cols] <- scale(test_data[, numeric_cols], center = train_means, scale = train_sds)

#checking results
summary(train_data[, numeric_cols])

```


Additional 11th-hour cleaning 
Removing highly correlated variables 
```{r}
#remove columns that are all NA
train_data <- train_data %>%
  select(where(~ !all(is.na(.))))

#define numeric predictors
numeric_cols <- train_data %>%
  select(where(is.numeric)) %>%
  select(-abundance, -relative_abundance, -log_abundance) %>%
  colnames()

#keeping  only the columns with no NA
train_numeric <- train_data[, numeric_cols] %>%
  select(where(~ all(!is.na(.))))

#correlation matrix
cor_matrix <- cor(train_numeric, use = "complete.obs")

#remove highly correlated variables
high_corr <- findCorrelation(cor_matrix, cutoff = 0.9)

numeric_cols_reduced <- colnames(train_numeric)[-high_corr]

#final dataset
train_data <- train_data %>%
  select(all_of(numeric_cols_reduced), log_abundance)




#Removing extreme outliers 
#WHY: so that extreme outliers don't impact my models 
abundance_data <- abundance_data %>%
  filter(abundance < quantile(abundance, 0.99))

```


```{r}
#looking at all my data together 
str(train_data)
summary(train_data)

#View(abundance_data)
View(train_data)
#View(test_data)
```


