---

title: "MapleModeling"
author: "Trinity Tobin"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
```{r}
#library bank
library(caret)
library(lmtest)
library(glmnet)

```



##DATA DICTIONARY 

Climate Variables (WorldClim)
Name	     Description	                                    Type
------------------------------------------------------------------------------------
BIO1	Annual mean temperature	                              Continuous
BIO4	Temperature seasonality	                              Continuous
BIO5	Max Temperature of Warmest Month	                    Continuous
BIO12	Annual Precipitation	                                Continuous
BIO15	Precipitation Seasonality	                            Continuous
BIO16	Precipitation of Wettest Quarter	                    Continuous

Species Occurrence Variables (GBIF)
Name	     Description	                                    Type
------------------------------------------------------------------------------------
species	   Scientific or common name of observed organism 
           (Acer saccharum, sugar maple)	                  Categorical
latitude	 Geographic coordinate (north-south )	            Continuous
longitude	 Geographic coordinate (east-west)	              Continuous
year	     Year the species occurrence was recorded	        Continuous (Discrete)


Variables of Significance (Variables Created From Feature Engineering)
Name	                 Description	                                     Type
------------------------------------------------------------------------------------
Longitude_band    Longitude grouped into 5-degree bins 	                 Numeric
Year              The year of the recorded maple observations	           Numeric
Year_centered     The recorded year centered by the mean year 	         Numeric
decimalLatitude   Mean latitude within each longitude band and year	     Numeric
decimalLongitude  Mean longitude within each longitude band and year	   Numeric
elevation         Mean elevation within each group	                     Numeric
Temp_range        Difference between BIO5 and BIO6	                     Numeric
Precip_range      Difference between BIO16 and BIO14	                   Numeric
Lag_abundance     Abundance from the previous year within the            Numeric
                  same longitude band	
Lag_log_abundance Log-transformed abundance from the previous year	     Numeric
Abundance_change  Change in abundance from the previous year	           Numeric
Density           Abundance divided by number of occurrences             Numeric
Log_abundance     Log-transformed abundance (log1p(abundance))	         Numeric



```{r cars}
#loading in the data and objects from my EDA and preprocessing 
load("Maple_preprocessing_workspace.RData")
```

```{r}
#checking that all my objects made it in 
ls()
str(train_data)
dim(train_data)
head(train_data)
colSums(is.na(train_data))
```



BASELINE MODEL #1: Linear Regression

WHY: A baseline linear regression model will allow me to:
        1. analyze the linear relationships between my target variable and my predictor variables
        2. serve as a benchnark model before moving on to more complex ones. 
        
```{r}
#LINEAR REGRESSION MODEL

#fitting the  model
lm_model <- lm(log_abundance ~ ., data = train_data)
summary(lm_model)


#cross-validation
set.seed(123)

train_control <- trainControl(
  method = "cv",
  number = 5
)

lm_cv <- train(
  log_abundance ~ .,
  data = train_data,
  method = "lm",
  trControl = train_control
)

lm_cv

#predictions
lm_train_pred <- predict(lm_model, newdata = train_data)
lm_test_pred  <- predict(lm_model, newdata = test_data)

#evaluation metrics
rmse <- function(actual, predicted) {
  sqrt(mean((actual - predicted)^2))
}

r2 <- function(actual, predicted) {
  cor(actual, predicted)^2
}

#training performance
lm_train_rmse <- rmse(train_data$log_abundance, lm_train_pred)
lm_train_r2   <- r2(train_data$log_abundance, lm_train_pred)

#testing performance
lm_test_rmse <- rmse(test_data$log_abundance, lm_test_pred)
lm_test_r2   <- r2(test_data$log_abundance, lm_test_pred)

#looking at my results 
lm_train_rmse
lm_train_r2
lm_test_rmse
lm_test_r2



#Residuals vs fitted plot: checks linearity and equal variance
#Q-Q plot: checks normality of residuals
#Scale-location plot: checks homoscedasticity
plot(lm_model)

#durbin watson
dwtest(lm_model)
```








BASELINE MODEL #2: Ridge Regression

WHY: Ridge regression is a regularized linear model that will 
        1. take into account the multicollinearity present in my data set (especially so among the BIO variables)
        2. as something to compare to the linear model

```{r}
#RIDGE REGRESSION MODEL

#making sure that the train and test have same columns
common_cols <- intersect(colnames(train_data), colnames(test_data))

train_ridge <- train_data[, common_cols]
test_ridge  <- test_data[, common_cols]

#create matrices
x_train <- model.matrix(log_abundance ~ ., data = train_ridge)[, -1]
y_train <- train_ridge$log_abundance

x_test <- model.matrix(log_abundance ~ ., data = test_ridge)[, -1]
y_test <- test_ridge$log_abundance

#force test matrix to match train matrix columns
missing_cols <- setdiff(colnames(x_train), colnames(x_test))

for (col in missing_cols) {
  x_test <- cbind(x_test, 0)
  colnames(x_test)[ncol(x_test)] <- col
}

extra_cols <- setdiff(colnames(x_test), colnames(x_train))

if (length(extra_cols) > 0) {
  x_test <- x_test[, !(colnames(x_test) %in% extra_cols)]
}

x_test <- x_test[, colnames(x_train)]

#cross-validated ridge regression
set.seed(123)

ridge_cv <- cv.glmnet(
  x_train,
  y_train,
  alpha = 0,
  nfolds = 5
)

best_lambda <- ridge_cv$lambda.min
best_lambda

#plotting the cross-validation
plot(ridge_cv)

#final ridge model
ridge_model <- glmnet(
  x_train,
  y_train,
  alpha = 0,
  lambda = best_lambda
)

#predictions
ridge_train_pred <- predict(ridge_model, s = best_lambda, newx = x_train)
ridge_test_pred  <- predict(ridge_model, s = best_lambda, newx = x_test)

#metrics
rmse <- function(actual, predicted) {
  sqrt(mean((actual - predicted)^2))
}

r2 <- function(actual, predicted) {
  cor(actual, predicted)^2
}

ridge_train_rmse <- rmse(y_train, ridge_train_pred)
ridge_train_r2   <- r2(y_train, ridge_train_pred)

ridge_test_rmse <- rmse(y_test, ridge_test_pred)
ridge_test_r2   <- r2(y_test, ridge_test_pred)

ridge_train_rmse
ridge_train_r2
ridge_test_rmse
ridge_test_r2

#testing assumptions with residuals plotting
ridge_residuals <- y_train - ridge_train_pred

ridge_diagnostics <- data.frame(
  fitted = ridge_train_pred,
  residuals = ridge_residuals
)


#qqplot to test normalcy (distribution)
qqnorm(ridge_residuals)
qqline(ridge_residuals, col = "red")


#durbin-watson
ridge_lm_for_dw <- lm(ridge_residuals ~ ridge_train_pred)
dwtest(ridge_lm_for_dw)

```


FUTURE: Random Forest
        Lasso





