---
title: "MapleModeling"
author: "Trinity Tobin"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
```{r}
#library bank
library(caret)
library(lmtest)
library(glmnet)
library(randomForest)
library(dplyr)
library(sf)
library(blockCV)

```



##DATA DICTIONARY 

Climate Variables (WorldClim)
Name	     Description	                                    Type
------------------------------------------------------------------------------------
BIO1	Annual mean temperature	                              Continuous
BIO4	Temperature seasonality	                              Continuous
BIO5	Max Temperature of Warmest Month	                    Continuous
BIO12	Annual Precipitation	                                Continuous
BIO15	Precipitation Seasonality	                            Continuous
BIO16	Precipitation of Wettest Quarter	                    Continuous

Species Occurrence Variables (GBIF)
Name	     Description	                                    Type
------------------------------------------------------------------------------------
species	   Scientific or common name of observed organism 
           (Acer saccharum, sugar maple)	                  Categorical
latitude	 Geographic coordinate (north-south )	            Continuous
longitude	 Geographic coordinate (east-west)	              Continuous
year	     Year the species occurrence was recorded	        Continuous (Discrete)


Variables of Significance (Variables Created From Feature Engineering)
Name	                 Description	                                     Type
------------------------------------------------------------------------------------
Longitude_band    Longitude grouped into 5-degree bins 	                 Numeric
Year              The year of the recorded maple observations	           Numeric
Year_centered     The recorded year centered by the mean year 	         Numeric
decimalLatitude   Mean latitude within each longitude band and year	     Numeric
decimalLongitude  Mean longitude within each longitude band and year	   Numeric
elevation         Mean elevation within each group	                     Numeric
Temp_range        Difference between BIO5 and BIO6	                     Numeric
Precip_range      Difference between BIO16 and BIO14	                   Numeric
Lag_abundance     Abundance from the previous year within the            Numeric
                  same longitude band	
Lag_log_abundance Log-transformed abundance from the previous year	     Numeric
Abundance_change  Change in abundance from the previous year	           Numeric
Density           Abundance divided by number of occurrences             Numeric
Log_abundance     Log-transformed abundance (log1p(abundance))	         Numeric



```{r cars}
#loading in the data and objects from my EDA and preprocessing 
load("Maple_preprocessing_workspace.RData")
```

```{r}
#checking that all my objects made it in 
ls()
str(train_data)
dim(train_data)
head(train_data)
colSums(is.na(train_data))
```



BASELINE MODEL #1: Linear Regression

WHY: A baseline linear regression model will allow me to:
        1. analyze the linear relationships between my target variable and my predictor variables
        2. serve as a benchnark model before moving on to more complex ones. 
        
```{r}
#LINEAR REGRESSION MODEL

#fitting the  model
lm_model <- lm(log_abundance ~ ., data = train_data)
summary(lm_model)


#cross-validation
set.seed(123)

train_control <- trainControl(
  method = "cv",
  number = 5
)

lm_cv <- train(
  log_abundance ~ .,
  data = train_data,
  method = "lm",
  trControl = train_control
)

lm_cv

#predictions
lm_train_pred <- predict(lm_model, newdata = train_data)
lm_test_pred  <- predict(lm_model, newdata = test_data)

#evaluation metrics
rmse <- function(actual, predicted) {
  sqrt(mean((actual - predicted)^2))
}

r2 <- function(actual, predicted) {
  cor(actual, predicted)^2
}

#training performance
lm_train_rmse <- rmse(train_data$log_abundance, lm_train_pred)
lm_train_r2   <- r2(train_data$log_abundance, lm_train_pred)

#testing performance
lm_test_rmse <- rmse(test_data$log_abundance, lm_test_pred)
lm_test_r2   <- r2(test_data$log_abundance, lm_test_pred)

#looking at my results 
lm_train_rmse
lm_train_r2
lm_test_rmse
lm_test_r2



#Residuals vs fitted plot: checks linearity and equal variance
#Q-Q plot: checks normality of residuals
#Scale-location plot: checks homoscedasticity
plot(lm_model)

#durbin watson
dwtest(lm_model)
```








BASELINE MODEL #2: Ridge Regression

WHY: Ridge regression is a regularized linear model that will 
        1. take into account the multicollinearity present in my data set (especially so among the BIO variables)
        2. as something to compare to the linear model

```{r}
#RIDGE REGRESSION MODEL

#making sure that the train and test have same columns
common_cols <- intersect(colnames(train_data), colnames(test_data))

train_ridge <- train_data[, common_cols]
test_ridge  <- test_data[, common_cols]

#create matrices
x_train <- model.matrix(log_abundance ~ ., data = train_ridge)[, -1]
y_train <- train_ridge$log_abundance

x_test <- model.matrix(log_abundance ~ ., data = test_ridge)[, -1]
y_test <- test_ridge$log_abundance

#force test matrix to match train matrix columns
missing_cols <- setdiff(colnames(x_train), colnames(x_test))

for (col in missing_cols) {
  x_test <- cbind(x_test, 0)
  colnames(x_test)[ncol(x_test)] <- col
}

extra_cols <- setdiff(colnames(x_test), colnames(x_train))

if (length(extra_cols) > 0) {
  x_test <- x_test[, !(colnames(x_test) %in% extra_cols)]
}

x_test <- x_test[, colnames(x_train)]

#cross-validated ridge regression
set.seed(123)

ridge_cv <- cv.glmnet(
  x_train,
  y_train,
  alpha = 0,
  nfolds = 5
)

best_lambda <- ridge_cv$lambda.min
best_lambda

#plotting the cross-validation
plot(ridge_cv)

#final ridge model
ridge_model <- glmnet(
  x_train,
  y_train,
  alpha = 0,
  lambda = best_lambda
)

#predictions
ridge_train_pred <- predict(ridge_model, s = best_lambda, newx = x_train)
ridge_test_pred  <- predict(ridge_model, s = best_lambda, newx = x_test)

#metrics
rmse <- function(actual, predicted) {
  sqrt(mean((actual - predicted)^2))
}

r2 <- function(actual, predicted) {
  cor(actual, predicted)^2
}

ridge_train_rmse <- rmse(y_train, ridge_train_pred)
ridge_train_r2   <- r2(y_train, ridge_train_pred)

ridge_test_rmse <- rmse(y_test, ridge_test_pred)
ridge_test_r2   <- r2(y_test, ridge_test_pred)

ridge_train_rmse
ridge_train_r2
ridge_test_rmse
ridge_test_r2

#testing assumptions with residuals plotting
ridge_residuals <- y_train - ridge_train_pred

ridge_diagnostics <- data.frame(
  fitted = ridge_train_pred,
  residuals = ridge_residuals
)


#qqplot to test normalcy (distribution)
qqnorm(ridge_residuals)
qqline(ridge_residuals, col = "red")


#durbin-watson
ridge_lm_for_dw <- lm(ridge_residuals ~ ridge_train_pred)
dwtest(ridge_lm_for_dw)

```

MODEL #3: Lasso  
WHY: Lasso is a regularized linear model that will 
        1. shrinks coefficients exactly to zero 
        2. also looks at multicollinearity 
        3. simpler than  the ridge regression (easier to interpret)
        
```{r}
rmse <- function(actual, predicted) {
  actual <- as.numeric(actual)
  predicted <- as.numeric(predicted)
  sqrt(mean((actual - predicted)^2))
}

r2 <- function(actual, predicted) {
  actual <- as.numeric(actual)
  predicted <- as.numeric(predicted)
  cor(actual, predicted)^2
}

#target variables
y_train <- train_data$log_abundance
y_test  <- test_data$log_abundance

#predictor data frames
train_x_df <- train_data %>%
  select(-log_abundance)

test_x_df <- test_data %>%
  select(-log_abundance)

#keep only columns that appear in both train and test
common_cols <- intersect(colnames(train_x_df), colnames(test_x_df))

train_x_df <- train_x_df[, common_cols]
test_x_df  <- test_x_df[, common_cols]

#create model matrices using the same columns
x_train <- model.matrix(~ ., data = train_x_df)[, -1]
x_test  <- model.matrix(~ ., data = test_x_df)[, -1]

#force test matrix to match train matrix exactly
missing_cols <- setdiff(colnames(x_train), colnames(x_test))

if (length(missing_cols) > 0) {
  missing_matrix <- matrix(
    0,
    nrow = nrow(x_test),
    ncol = length(missing_cols)
  )
  colnames(missing_matrix) <- missing_cols
  x_test <- cbind(x_test, missing_matrix)
}

extra_cols <- setdiff(colnames(x_test), colnames(x_train))

if (length(extra_cols) > 0) {
  x_test <- x_test[, !(colnames(x_test) %in% extra_cols)]
}

x_test <- x_test[, colnames(x_train)]

#check dimensions before modeling
dim(x_train)
dim(x_test)
length(y_train)
length(y_test)

#cross-validated Lasso
set.seed(123)

lasso_cv <- cv.glmnet(
  x_train,
  y_train,
  alpha = 1,
  nfolds = 5
)

best_lambda_lasso <- lasso_cv$lambda.min
best_lambda_lasso

plot(lasso_cv)

#fit final Lasso model
lasso_model <- glmnet(
  x_train,
  y_train,
  alpha = 1,
  lambda = best_lambda_lasso
)

#predictions
lasso_train_pred <- as.numeric(predict(lasso_model, s = best_lambda_lasso, newx = x_train))
lasso_test_pred  <- as.numeric(predict(lasso_model, s = best_lambda_lasso, newx = x_test))

#check prediction lengths
length(lasso_train_pred)
length(lasso_test_pred)
length(y_train)
length(y_test)

#metrics
lasso_train_rmse <- rmse(y_train, lasso_train_pred)
lasso_train_r2   <- r2(y_train, lasso_train_pred)

lasso_test_rmse <- rmse(y_test, lasso_test_pred)
lasso_test_r2   <- r2(y_test, lasso_test_pred)

lasso_train_rmse
lasso_train_r2
lasso_test_rmse
lasso_test_r2

#selected variables
coef(lasso_model, s = best_lambda_lasso)
```


^^ I did this before reading all of the comments on the Baseline report but will be disregarding for further analysis 


MODEL #4: Random Forest   
WHY: Random forest is a non-linear model that will 
        1. looks at nonlinear relationships and interactions
        2. good for ecology
        3. will let me see which features are most important 
        4. robust aginst overfitting (because of the many trees)
        
```{r}
rmse <- function(actual, predicted) {
  sqrt(mean((actual - predicted)^2))
}

r2 <- function(actual, predicted) {
  cor(actual, predicted)^2
}

set.seed(123)

rf_model <- randomForest(
  log_abundance ~ .,
  data = train_data,
  ntree = 500,
  mtry = floor(sqrt(ncol(train_data) - 1)),
  importance = TRUE
)

rf_model

#predictions
rf_train_pred <- predict(rf_model, newdata = train_data)
rf_test_pred  <- predict(rf_model, newdata = test_data)

#metrics
rf_train_rmse <- rmse(train_data$log_abundance, rf_train_pred)
rf_train_r2   <- r2(train_data$log_abundance, rf_train_pred)

rf_test_rmse <- rmse(test_data$log_abundance, rf_test_pred)
rf_test_r2   <- r2(test_data$log_abundance, rf_test_pred)

rf_train_rmse
rf_train_r2
rf_test_rmse
rf_test_r2

#variable importance
importance(rf_model)
varImpPlot(rf_model)
```


## ADRESSING ISSUES/CONCERNS WITH MY PROJECT (BASED ON FEEDBACK)


1. I know you wouldn't have seen my EDA comments about changing abundance to "occurrence" or something similar yet, but here's another reminder. :) IMPORTANT: My biggest concern is over replacing missing individualCount values with 1. Why not 0? Or even just doing sums that ignore NA? This needs much stronger justification because currently it assumes absence of information equals a single occurrence, which will very likely bias abundance estimates.


PROBLEM: replacing missing values in individualCount with 1 
WHY: assuming no info is an occurrence

SOLUTION: changing the variable from "abundance" to "occurrence_count"
WHY: this will elminate the assumption of an occurrence and serve as a more accurate quantification based on the true occurrences
```{r}


#making my response variable based on occurrences 
# Each GBIF row is treated as one occurrence record
sugar_climate_clean <- sugar_climate_clean %>%
  mutate(occurrence_record = 1)

#grouping by longitude_band-year level
occurrence_data <- sugar_climate_clean %>%
  group_by(longitude_band, year) %>%
  summarise(
    occurrence_count = n(),
    reported_individual_count = sum(individualCount, na.rm = TRUE),
    records_with_individual_count = sum(!is.na(individualCount)),
    decimalLatitude = mean(decimalLatitude, na.rm = TRUE),
    decimalLongitude = mean(decimalLongitude, na.rm = TRUE),
    elevation = mean(elevation, na.rm = TRUE),
    across(starts_with("BIO"), ~ mean(.x, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(
    log_occurrence_count = log1p(occurrence_count)
  ) %>%
  group_by(year) %>%
  mutate(
    relative_occurrence = occurrence_count / sum(occurrence_count, na.rm = TRUE)
  ) %>%
  ungroup()

#creating temporal lags and derived variables
occurrence_data <- occurrence_data %>%
  arrange(longitude_band, year) %>%
  group_by(longitude_band) %>%
  mutate(
    lag_occurrence_count = lag(occurrence_count, 1),
    lag_log_occurrence_count = lag(log_occurrence_count, 1),
    lag_relative_occurrence = lag(relative_occurrence, 1),
    occurrence_change = occurrence_count - lag_occurrence_count
  ) %>%
  ungroup() %>%
  mutate(
    temp_range = BIO5 - BIO6,
    precip_range = BIO16 - BIO14,
    precip_seasonality = BIO15,
    year_centered = year - mean(year, na.rm = TRUE)
  ) %>%
  filter(!is.na(lag_occurrence_count))

head(occurrence_data)
colSums(is.na(occurrence_data))



#new train and test data splits (temporal split)
train_data <- occurrence_data %>%
  filter(year < 2015)

test_data <- occurrence_data %>%
  filter(year >= 2015)




#adding the spatial split 
set.seed(123)

#get unique longitude bands
unique_bands <- unique(occurrence_data$longitude_band)

#randomly assign 80% of longitude bands to training
train_bands <- sample(
  unique_bands,
  size = floor(0.8 * length(unique_bands))
)

#split data by longitude band
train_data <- occurrence_data %>%
  filter(longitude_band %in% train_bands)

test_data <- occurrence_data %>%
  filter(!longitude_band %in% train_bands)

#check split
nrow(train_data)
nrow(test_data)

length(unique(train_data$longitude_band))
length(unique(test_data$longitude_band))

unique(train_data$longitude_band)
unique(test_data$longitude_band)

```
```{r}
#the new columns still have missing values from the lagging of the variables
#we fix that here with the same median imputation that we did before 

#check missing values
#colSums(is.na(train_data))
#colSums(is.na(test_data))

#remove rows missing the target variable
train_data <- train_data %>%
  filter(!is.na(log_occurrence_count))

test_data <- test_data %>%
  filter(!is.na(log_occurrence_count))

#identify numeric predictor columns
numeric_cols <- train_data %>%
  select(where(is.numeric)) %>%
  select(-log_occurrence_count) %>%
  colnames()

#calculate medians from training data only
train_medians <- sapply(train_data[, numeric_cols], median, na.rm = TRUE)

#impute train and test using training medians
for (col in numeric_cols) {
  train_data[[col]][is.na(train_data[[col]])] <- train_medians[[col]]
  test_data[[col]][is.na(test_data[[col]])] <- train_medians[[col]]
}

#check again
colSums(is.na(train_data))        #left only one so it was easier to see 
#colSums(is.na(test_data))
```
```{r}
#POSSIBLE LEAKAGE
#removing variables with the target to avoid leakage 
rf_train <- train_data %>%
  select(-occurrence_count, -relative_occurrence)

rf_test <- test_data %>%
  select(-occurrence_count, -relative_occurrence)
```


## NEW RANDOM FOREST CODE WITH MY NEW TARGET VARIABLE ("occurrence_count")
```{r}
rmse <- function(actual, predicted) {
  sqrt(mean((actual - predicted)^2))
}

r2 <- function(actual, predicted) {
  cor(actual, predicted)^2
}

set.seed(123)

rf_model <- randomForest(
  log_occurrence_count ~ .,
  data = rf_train,
  ntree = 500,
  mtry = floor(sqrt(ncol(rf_train) - 1)),
  importance = TRUE
)

rf_train_pred <- predict(rf_model, newdata = rf_train)
rf_test_pred  <- predict(rf_model, newdata = rf_test)

rf_train_rmse <- rmse(rf_train$log_occurrence_count, rf_train_pred)
rf_train_r2   <- r2(rf_train$log_occurrence_count, rf_train_pred)

rf_test_rmse <- rmse(rf_test$log_occurrence_count, rf_test_pred)
rf_test_r2   <- r2(rf_test$log_occurrence_count, rf_test_pred)

rf_train_rmse
rf_train_r2
rf_test_rmse
rf_test_r2

importance(rf_model)
varImpPlot(rf_model)
```





##MY NEXT STEP:
   1. CHOOSING MY BEST MODEL 
   2. CLEANING LEAKAGE IN THE MODEL (MAIN FORMS COME FROM SIMILAR VARIABLES)
   3. HYPER PARAMETER TUNING (FOR SAID MODEL)


#At this point, I have 3 models: linear regression, ridge regression, and random forest. The mdoel that I am going to move forward with for my final deliverable is 
     
  RANDOM FOREST
  
WHY:     1. great for nonlinearity
         2. will be able to capture interactions in the variables
         3. there are not the same assumptions as the linear tests (ec. normal, linear, constant variance)
         4. the trees can help against overfitting
         5. the above reasons make it well suited for ecological data 
         
```{r}
#Addressing leakage in multiple ways 

#1. removing variables with the target variable (line 551 above, as well)

set.seed(123)

#removing variables associated with leakage
#because my target variable is occurrence_count, any related variables will result in possible leakage, so for this model, I removed closely corrleated variables 
leakage_vars <- c(
  "reported_individual_count",
  "records_with_individual_count",
  "occurrence_change",
  "lag_occurrence_count",
  "lag_log_occurrence_count",
  "lag_relative_occurrence"
)

rf_train_noleak <- rf_train %>%
  select(-any_of(leakage_vars))

rf_test_noleak <- rf_test %>%
  select(-any_of(leakage_vars))

#adding spatial cross-validation
    #converting my training data to spatial object
rf_train_sf <- st_as_sf(
  rf_train_noleak,
  coords = c("decimalLongitude", "decimalLatitude"),
  crs = 4326,
  remove = FALSE
)

    #creating the spatial cross-validation folds 
spatial_folds <- cv_spatial(
  x = rf_train_sf,
  column = "log_occurrence_count",
  k = 5,
  size = 500000,
  selection = "random",
  iteration = 100,
  progress = TRUE
)
    
    #changing format of folds so it works for the model 
train_index <- spatial_folds$folds_list %>%
  lapply(function(x) x[[1]])

test_index <- spatial_folds$folds_list %>%
  lapply(function(x) x[[2]])


   #the code to actuallly create the spatial folds
spatial_cv_control <- trainControl(
  method = "cv",
  index = train_index,
  indexOut = test_index,
  verboseIter = TRUE
)

#looking at how removing the variables affected the RMSE and R2
rmse <- function(actual, predicted) {
  sqrt(mean((actual - predicted)^2, na.rm = TRUE))
}

r2 <- function(actual, predicted) {
  cor(actual, predicted, use = "complete.obs")^2
}

#spatial cross-validation random forest 
tune_grid <- expand.grid(
  mtry = c(3, 5, 7, 10, 15)
)

rf_spatial_cv_model <- train(
  log_occurrence_count ~ .,
  data = rf_train_noleak,
  method = "rf",
  trControl = spatial_cv_control,  # 🔥 THIS IS WHERE CV IS USED
  tuneGrid = tune_grid,
  ntree = 500,
  importance = TRUE
)

#finsing the best parameter
rf_spatial_cv_model$bestTune

#cross validation results
rf_spatial_cv_model$results
```

```{r}
#HYPERPARAMETER TUNING WITH SPATIAL CROSS-VALIDATION

set.seed(123)

#defining a tuning grid (mtry only for caret RF)


tune_grid <- expand.grid(
  mtry = c(3, 5, 7, 10, 15)
)


#training the model using spatial CV folds
rf_spatial_cv_model <- train(
  log_occurrence_count ~ .,
  data = rf_train_noleak,
  method = "rf",
  trControl = spatial_cv_control,
  tuneGrid = tune_grid,
  ntree = 500,
  importance = TRUE
)

#looking at the best hyperparameter + tuning results
rf_spatial_cv_model$bestTune
rf_spatial_cv_model$results


#evaluating the cross-validation on held-out test set
rf_spatial_test_pred <- predict(
  rf_spatial_cv_model,
  newdata = rf_test_noleak
)

rmse <- function(actual, predicted) {
  sqrt(mean((actual - predicted)^2, na.rm = TRUE))
}

r2 <- function(actual, predicted) {
  cor(actual, predicted, use = "complete.obs")^2
}

rf_spatial_test_rmse <- rmse(
  rf_test_noleak$log_occurrence_count,
  rf_spatial_test_pred
)

rf_spatial_test_r2 <- r2(
  rf_test_noleak$log_occurrence_count,
  rf_spatial_test_pred
)

rf_spatial_test_rmse
rf_spatial_test_r2

```


```{r}
##FINAL TUNED NO-LEAKAGE RANDOM FOREST MODEL

set.seed(123)
rf_spatial_test_pred <- predict(
  rf_spatial_cv_model,
  newdata = rf_test_noleak
)

spatial_test_rmse <- rmse(
  rf_test_noleak$log_occurrence_count,
  rf_spatial_test_pred
)

spatial_test_r2 <- r2(
  rf_test_noleak$log_occurrence_count,
  rf_spatial_test_pred
)

spatial_test_rmse
spatial_test_r2

#variable importance
varImp(rf_spatial_cv_model)
plot(varImp(rf_spatial_cv_model))
```





boop







|
v later issues 
--------------------------------------------------------------------------------------------

2. Your formal preprocessing pipeline is still incomplete/unclear. 
IMPORTANT: You have some solid steps here but I need a little more clarity in a few places. You did a great job with the creation of the spatial (longitude) bands! Also you did well with making your temporal lags. However, 

   1. it's unclear if you applied that to both train and test or just train or -- more appropriately -- spatial bands to train/test and temporal lags ONLY to train. Also, 
   2. clarify: did you do a temporal split? If so, what were the year cutoffs? 
   3. And is an 80/20 split appropriate for the dimensionality of your data? Also, 
   4. your order of operations is unclear, specifically whether imputation and scaling were also performed using training data only. These kinds of things can be cleared up quickly with just an extra clarifying sentence or two. 
   5. HOWEVER - given the temporal and spatial structure (lags, longitudinal bands), there is also a risk of leakage that is not addressed. So, in your Final Deliverable, aim to provide a clearer, step-by-step pipeline that explicitly handles these issues. How are you being careful not to introduce possible leakage? And are you creating any opportunities for bias your reader should be aware of?


PROBLEM: general cleanup and clarification of the pipeline (clean through text writing in both the deliverable and the RMarkdown)
WHY: it's unclear (straight up)

SOLUTION:


   
   
   